Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-10 paper/ch_potentials.tex
changes to A* text

On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
A bidirectional variant of A* therefore appears also desirable.
However, as shown in~\cite{gh-cspas-05}, this is not as simple as it seems.

The straightforward idea would be to use two heuristics.
The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$ where $\overrightarrow{h}_t(x)$ estimates the distance from $x$ to $t$.
Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$ where $\overleftarrow{h}_s(x)$ is an estimate of the distance from $s$ to $x$.
% We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a better stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined heuristic $h_{\overline{st}}$, which is good for both directions.
Both searches run on the same reduced graph which enables an efficient stopping criterion.
The reduced graph of the average potential is the graph with the average weights of the reduced graphs of $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
The heuristic function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$.
The backward search uses $-h_{\overline{st}}(x)$.
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
Unfortunately, it has two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
This makes it more expensive in terms of running time.
% Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
Second, the better stopping criterion comes at the cost of worse estimates for each direction on its own.
The average potential estimates $\dist(x,t)$ less accurately than $\overrightarrow{h}_t(x)$ and $\dist(s,x)$ less accurately than $\overleftarrow{h}_s(x)$.
The second downside can be partially mitigated through pruning with the original heuristics.
When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(xy) + \overrightarrow{h}_s(y) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.
The pruning for the backward search works analogously.

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.
Recall that queues are ordered by $D[x] + h(x)$.
If a node $x$ was not yet settled by backward search then $\dist(x,t) + \overleftarrow{h}_s(x) \geq \overleftarrow{k}$, and equivalently $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
The pruning for the backwards search works analogously.
As the pruning becomes tighter as the searches progress, it compensates for the worse stopping criterion.
Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
As our experiments show, this makes the simpler symmetric approach competitive to the average potential.

>>>>

On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
A bidirectional variant of A* therefore is therefore also desirable.
However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straight forward.
We revisit bidirectional A* and propose an alternative approach.
Our experiments show that it is competitive with the solution described by~\cite{gh-cspas-05}.

The straightforward idea is to use two heuristics $\overrightarrow{h}_t(x)$ and $\overleftarrow{h}_s(x)$.
The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$, where $\overrightarrow{h}_t(x)$ estimates the distance $\dist(x,t)$ from $x$ to $t$.
Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$, where $\overleftarrow{h}_s(x)$ is an estimate of the distance $\dist(s,x)$ from $s$ to $x$.
% We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
This is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} therefore refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a bidirectional stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
The forward search uses $h_{\overline{st}}(x)$ as heuristic.
The backward search uses $-h_{\overline{st}}(x)$.
This results in both searches running on the same reduced graph.
The weights of the combined reduced graph corresponds to the averaged weights of the individual reduced graphs.\todo{Ist diese Aussage korrekt?}
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
To the best of our knowledge, using this stopping criterion yields the fastest known bidirectional A* variant.
Unfortunately, it has two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
Evaluating the average heuristic is therefore slower than evaluating just one the composing heuristics.
Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
$h_{\overline{st}}(x)$ is a worse estimate for $\dist(x,t)$ than $\overrightarrow{h}_t$.
Similarly, $-h_{\overline{st}}(x)$ is a worse estimate for $\dist(s,x)$ than $\overleftarrow{h}_s$.
The second downside can be partially mitigated through pruning with the composing heuristics.
When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(x,y) + \overrightarrow{h}_s(y) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.\todo{does not have to be inserted? needs not klingt komisch... aber English ist nicht meine Stärke. :) }
The pruning rule for the backward search is analogous.

\todo{$w(xy)$ vs $w(x,y)$}
To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We do not use the average heuristic and use the unidirectional stopping criteria.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
Let $xy$ be an edge that we relax in the forward search.
Before pushing $y$ into the queue, we apply the pruning rule.
If we can prove that every path using $xy$ is at least as long as the shortest known path length $\mu$, then we do not have to push $y$.
We therefore want to obtain a lower bound for $\dist(s,x)+w(x,y)+\dist(y,t)$.
As $x$ was settled, $\overrightarrow{D}[x]$ contains the shortest path length $\dist(s,x)$, i.e., $\dist(s,x)=\overrightarrow{D}[x]$.
$w(x,y)$ is also exactly known as it's just an edge weight.
It remains to lower bound $\dist(y,t)$.
Nodes are removed from the backward queue ordered by $\overrightarrow{h}(y) + \dist(y,t)$.
\todo{$\ge$ vs $\geq$ und $\le$ vs $\leq$, macht das einen Unterschied?}
If $y$ was not yet removed from the backward queue, we know that $\overrightarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
This gives us the required lower bound $\dist(y,t) \geq \overleftarrow{k}-\overrightarrow{h}(y)$.
$y$ does not have to be pushed, if it was not settled by the backward search and $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{k}-\overrightarrow{h}(y)\ge \mu$.
This does not mean that $y$ will never be pushed into the forward queue.
It is possible that another edge $zy$ exists for which the pruning rule cannot be applied.

\todo{Latex mag overrightarrow nicht in todos... Ich packe die Frage deswegen inline in den Text.}
Kann man sinnvoll prunen wenn $y$ von der Rückwärtssuche gesettled wurde? Z.b. Kante $xy$ prunen wenn $x$ forward gestelled ist und $y$ backward gesettled ist und $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{D}[y]\ge \mu$ erfüllt ist.

\todo{Der folgende Paragraph ist die Vorgängerversion vom Paragraph davor. Wenn der neue Paragraph besser ist, dann kann dieser weg.}
We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.
Recall that queues are ordered by $D[x] + h(x)$.
If a node $x$ was not yet settled by backward search then $\dist(x,t) + \overleftarrow{h}_s(x) \geq \overleftarrow{k}$, and equivalently $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
The pruning for the backwards search works analogously.
As the pruning becomes tighter as the searches progress, it compensates for the worse stopping criterion.
Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
